A hierarchical model to evaluate pest treatments from prevalence and intensity data

### 2023-10-06 submitted-202310-favrot-pest.qmd
suprresing files not meant for control versioning, .gitignore update and others set-up for fresh restart

keywords: [bayesian model, epidemiology, hierarchical model, pest control, trial, survey]
doi: "to complete"
citation:
  type: article-journal
  container-title: "Computo"
  issued: "2023/01/25"
  url: https://computo.sfds.asso.fr/
  issn: "2824-7795"
google-scholar: true
bibliography: references.bib
github-user: computorg
repo: "To complete"
draft: false 
published: false
format:
  computo-pdf: default
header-includes:
  - \usepackage{lineno}
---

@@

```{r figure1, cache = TRUE, fig.width=12, fig.height=14}
#| echo: false
#| label: fig-one 
#| fig-cap: "Description of the dataset. **A** Number of observations according to the type of insecticide. **B** Examples of observed number of aphids averaged over the blocks (intensity) and number of infested beets out of ten (prevalence) averaged over the blocks, at different dates for two trials."

@@

```{r inference_example_real_data, cache = TRUE, results = 'hide'}
#| echo: true
#| eval: false
#| file: scripts/inference_example_real_data.R
```

@@

| Type of dataset | Description |
|---------|:-----|
| 100% Y - 0% Z      | Y observations available in the 32 trials   |
| 50% Y - 0% Z     | Y observations available in 17 trials and no Z  | 
| 0% Y - 50% Z       | Z observations available in 15 trials and no Y    | 
| 50% Y - 50% Z       | Y observations available in 17 trials and Z observations available in the other 15 trials   |

@@

- Draw values of $\beta_{0i},\ u_{ij}$ and $\epsilon_{ijkt}$ in their distributions for each trial, 3 treatments (+ the untreated control), 3 dates ($t$=0, 6, 12), and 4 blocks
- Calculate $\lambda_{ijkt}$ from @eq-model_lambda,
- Draw values of $W_{ijkts}$ in its Poisson distribution for 10 plants ($s$=1, ..., 10)
- Calculate $Y_{ijkt}$ and $Z_{ijkt}$ from the Ws for each trial, treatment, date, block.
- Generate the eight sub datasets corresponding to the scenarios defined in Table 3 (including all values of $W$, the generated values of $Y$ only, the generated values of $Z$ only, both $Y$ and $Z$ values but not $W$, the values of $Y$ in 50% of the trials, the values of $Z$ in 50% of the trials, the values of $Y$ in 50% of the trials and the values of $Z$ in the other 50%), 
- Fit the model (@eq-model_W - @eq-model_Z) to each of the sub datasets according to the procedure described above based on MCMC.

@@

where the true parameter values are set equal to the posterior means computed with the real dataset, and the parameter estimates ($\hat{\gamma}_{0j}$ and $\hat{\gamma}_{1j}$) are the posterior means computed for the jth treatment. For each trial number and each scenario defined in Table 3, the 974 values of $E_{\gamma}$ obtained for the 974 generated sub datasets are then averaged. The average values obtained for the eight scenarios are finally compared to determine the type of data leading to the most accurate estimated parameter values.

@@

```{r simulation-function, cache = TRUE}
#| echo: true
#| eval: true 
#| file: functions/simu_data.R 
```

@@

::: {.callout-note}
The following code presents the inference for one seed. As the computation time is growing fast with the number of trials and the number of sampled values, we set here the number of trials to 10 (I = 10) and the number of adaptations and iterations to 2000.

Our simulation results are obtained by running this code for I = 10, 20, 40, 80, for all seeds between 0 and 999, with a number of adaptations and iterations equal to 36000. We used the Migale server from INRAE to run these simulations.

@@

```{r simulations, results = 'hide', cache = TRUE}
#| echo: true
#| eval: false
#| file: scripts/simulations.R
```

@@

```{r figure2, cache = TRUE, out.width="100%", fig.height = 7, echo = FALSE}
#| label: fig-two
#| fig-cap: "Model convergence - Markov chain for the model parameters and treatment efficacies, in the scenarios '50% Y - 0% Z' (**A**),  '50% Y - 50% Z' (**B**), '100% Y - 0% Z' (**C**) and '0% Y - 50% Z' (**D**). The x-axis presents the iteration number and the y-axis presents the sampled value."

@@

```{r figure3, cache = TRUE, out.width="100%", fig.height = 7, echo = FALSE}
#| label: fig-three
#| fig-cap: "Model convergence - Gelman-Rubin statistics for the parameters of the model (@eq-model_W - @eq-model_Z) and for treatment efficacies (@eq-Efficacy), according to the scenarios 50% Y - 0% Z (**A**),  50% Y - 50% Z (**B**), 100% Y - 0% Z (**C**) and 0% Y - 50% Z (**D**). The x-axis presents the iteration number and the y-axis presents the Gelman-Rubin statistic."

@@

```{r table4, cache = TRUE}
#| echo: false
tab4 = cbind(summary(resY)$statistics, summary(resY)$quantiles) %>% 
    as.data.frame %>%
    rownames_to_column(var = "Parameter") %>% select(- c(4, 5, c(7 : 9))) %>%
    mutate(Parameter = recode(Parameter, "gamma0[1]" = "gamma0 - untreated", "gamma0[2]" = "gamma0 - Mavrik Jet",
                                         "gamma0[3]" = "gamma0 - Movento", "gamma0[4]" = "gamma0 - Teppeki",
                                          "gamma1[1]" = "gamma1 - untreated", "gamma1[2]" = "gamma1 - Mavrik Jet",
                                         "gamma1[3]" = "gamma1 - Movento", "gamma1[4]" = "gamma1 - Teppeki"  )) %>%
    filter(!grepl("Eff", Parameter)) %>%

@@

tab4 %>% 
  kbl(booktabs = TRUE, caption = "Summary of the posterior distributions obtained with the 100 Y scenario for the model parameters and treatment efficacies: posterior mean, standard deviation, 2.5 and 97.5 quantiles.") %>%
  kable_styling()
```

@@

```{r figure4, cache = TRUE, echo=FALSE, out.width = "100%", fig.height = 2.6}
#| label: fig-four
#| fig-cap: "Estimated treatment efficacies after 6 days (**A**) and after 12 days (**B**), with their credibility intervals. Colors correspond to the different scenarios."

@@

```{r table5, cache = TRUE}
#|echo: false
cbind(df_len6, df_len12 %>% select(- Data, - Insecticide)) %>% 
  kbl(booktabs = TRUE, caption = "Differences in the sizes of the 95 credibility intervals (CI) of the estimated treatment efficacies for the scenarios 50 Y / 50 Z, 50 Z and 100 Y, compared to 50 Y. The difference is given in percentage. A positive (negative) value indicates an increase (decrease) of the credibility interval size. The third column indicates differences for the efficacy at 6 days after pesticide spray, and the fourth column indicates the difference for the efficacy at 12 days.") %>% 
  kable_styling()
```

@@

```{r load_res_simulations, cache = TRUE, echo = FALSE}
df_col_simu = readRDS(file = "Files_for_code/Results/Simulations/df_col_simu.rds")
res_simu_sigma_1.87 = readRDS(file = "Files_for_code/Results/Simulations/res_simu_sigma_1.87.rds")

@@

```{r table6, cache = TRUE}
#|echo: false

@@

: Parameters used to generate virtual data.

@@

```{r figure5, cache = TRUE, out.width="100%", fig.height = 2.5, echo = FALSE}
#| label: fig-five
#| fig-cap: "Values of the $E_{\\gamma}$ (@eq-E_gamma) (**A**), $E_{Ef_6}$ (@eq-E_efficacy) (**B**) and $E_{Ef_{12}}$ (@eq-E_efficacy) (**C**) mean absolute error criteria for the '0% Y - 50% Z', '50% Y - 0% Z' and '50% Y - 50% Z' scenarios. The x-axis presents the number of trials and the y-axis presents the mean absolute error, averaged over the 974 simulated datasets. Different colors correspond to different scenarios."

@@

```{r figure6, cache = TRUE, out.width="100%", fig.height = 3, echo = FALSE}
#| label: fig-six
#| fig-cap: "Comparison of proportion cases where the best treatment is correctly identified in the '0% Y - 50% Z', '50% Y - 0% Z' and '50% Y - 50% Z' scenarios. The x-axis represents the number of trials and the y-axis represents the percentage of cases where the best treatment has been correctly identified at 6 days (A) and 12 days (B), over the 974 simulated datasets. Different colors correspond to different scenarios."

@@

```{r figure7, cache = TRUE, out.width="100%", fig.height = 2.5, echo = FALSE}
#| label: fig-seven
#| fig-cap: "Values of $E_{\\gamma}$ (@eq-E_gamma) (**A**), $E_{Ef_6}$ (@eq-E_efficacy) (**B**) and $E_{Ef_{12}}$ (@eq-E_efficacy) (**C**) for the '0% Y - 100% Z', '50% Y - 0% Z', '50% Y - 100% Z' and '50% Y - 50% Z' scenarios. The x-axis presents the number of trials and the y-axis presents the absolute error averaged over the 974 simulated datasets. Different colors correspond to different scenarios."

@@

```{r table7, cache = TRUE}
#|echo: false

@@

```{r figure8, cache = TRUE, out.width="100%", fig.height = 6, echo = FALSE}
#| label: fig-height
#| fig-cap: "Comparison of the '100% Y - 0% Z', '0% Y - 100% Z', '100% Y / 100% Z' and '100% W' scenarios according to the distribution of $Z$ and the number of trials, using the $E_{\\gamma}$ criterion (@eq-E_gamma). **A**, **B** and **C** correspond to different $Z$ distributions which are given by A.2, B.2 and C.2 (distribution for a number of trials equal to 40). A, B and C respectively correspond to $\\alpha_0 =$ -1, 1 and 2. The details of the simulation parameters are given in table 7. A1, B1 and C1 represent the absolute error $E_{\\gamma}$ averaged over the 974 simulated datasets as a function of the number of trials. Colors correspond to the different scenarios."

>>>>

keywords: [bayesian model, epidemiology, hierarchical model, pest control, trial, survey]
doi: "to complete"
citation:
  type: article-journal
  container-title: "Computo"
  issued: "2023/01/25"
  url: https://computo.sfds.asso.fr/
  issn: "2824-7795"
google-scholar: true
bibliography: references.bib
github-user: computorg
repo: "To complete"
draft: false 
published: false
format:
  computo-html: default
  computo-pdf: default
header-includes:
  - \usepackage{lineno}
---

@@

```{r figure1, cache = FALSE, fig.width=12, fig.height=14}
#| echo: false
#| label: fig-one 
#| fig-cap: "Description of the dataset. **A** Number of observations according to the type of insecticide. **B** Examples of observed number of aphids averaged over the blocks (intensity) and number of infested beets out of ten (prevalence) averaged over the blocks, at different dates for two trials."

@@

```{r inference_example_real_data, cache = FALSE, results = 'hide'}
#| echo: true
#| eval: true
#| file: scripts/inference_example_real_data.R
```

@@

| Type of dataset | Description |
|---------|:-----|
| 100% Y - 0% Z      | Y observations available in the 32 trials and no Z  |
| 50% Y - 0% Z     | Y observations available in 17 trials and no Z  | 
| 0% Y - 50% Z       | Z observations available in 15 trials and no Y    | 
| 50% Y - 50% Z       | Y observations available in 17 trials and Z observations available in the other 15 trials   |

@@

- Draw values of $\beta_{0i},\ u_{ij}$ and $\epsilon_{ijkt}$ in their distributions for each trial, 3 treatments (+ the untreated control), 3 dates ($t$=0, 6, 12), and 4 blocks
- Calculate $\lambda_{ijkt}$ from @eq-model_lambda,
- Draw values of $W_{ijkts}$ in its Poisson distribution for 10 plants ($s$=1, ..., 10)
- Calculate $Y_{ijkt}$ and $Z_{ijkt}$ from the Ws for each trial, treatment, date, block.
- Generate the eight data subsets corresponding to the scenarios defined in Table 3 (including all values of $W$, the generated values of $Y$ only, the generated values of $Z$ only, both $Y$ and $Z$ values but not $W$, the values of $Y$ in 50% of the trials, the values of $Z$ in 50% of the trials, the values of $Y$ in 50% of the trials and the values of $Z$ in the other 50%), 
- Fit the model (@eq-model_W - @eq-model_Z) to each of the data subsets according to the procedure described above based on MCMC.

@@

where the true parameter values are set equal to the posterior means computed with the real dataset, and the parameter estimates ($\hat{\gamma}_{0j}$ and $\hat{\gamma}_{1j}$) are the posterior means computed for the jth treatment. For each trial number and each scenario defined in Table 3, the 974 values of $E_{\gamma}$ obtained for the 974 generated data subsets are then averaged. The average values obtained for the eight scenarios are finally compared to determine the type of data leading to the most accurate estimated parameter values.

@@

```{r simulation-function, cache = FALSE}
#| echo: true
#| eval: true 
#| file: functions/simu_data.R 
```

@@

::: {.callout-note}
The following code presents the inference for one seed. As the computing time is growing fast with the number of trials and the number of sampled values, we set here the number of trials to 10 (I = 10) and the number of adaptations and iterations to 2000.

Our simulation results are obtained by running this code for I = 10, 20, 40, 80, for all seeds between 0 and 999, with a number of adaptations and iterations equal to 36000.

@@

```{r simulations, results = 'hide', cache = FALSE}
#| echo: true
#| eval: true
#| file: scripts/simulations.R
```

@@

```{r figure2, cache = FALSE, out.width="100%", fig.height = 7, echo = FALSE}
#| label: fig-two
#| fig-cap: "Model convergence - Markov chain for the model parameters and treatment efficacies, in the scenarios '50% Y - 0% Z' (**A**),  '50% Y - 50% Z' (**B**), '100% Y - 0% Z' (**C**) and '0% Y - 50% Z' (**D**). The x-axis presents the iteration number and the y-axis presents the sampled value."

@@

```{r figure3, cache = FALSE, out.width="100%", fig.height = 7, echo = FALSE}
#| label: fig-three
#| fig-cap: "Model convergence - Gelman-Rubin statistics for the parameters of the model (@eq-model_W - @eq-model_Z) and for treatment efficacies (@eq-Efficacy), according to the scenarios '50% Y - 0% Z' (**A**),  '50% Y - 50% Z' (**B**), '100% Y - 0% Z' (**C**) and '0% Y - 50% Z' (**D**). The x-axis presents the iteration number and the y-axis presents the Gelman-Rubin statistic."

@@

```{r table4, cache = FALSE}
#| echo: false
tab4 = cbind(summary(resY)$statistics, summary(resY)$quantiles) %>% 
    as.data.frame %>%
    rownames_to_column(var = "Parameter") %>% select(- c(4, 5, c(7 : 9))) %>%
    mutate(Parameter = recode(Parameter, "gamma0[1]" = "gamma0 - untreated", "gamma0[2]" = "gamma0 - Mavrik Jet",
                                         "gamma0[3]" = "gamma0 - Movento", "gamma0[4]" = "gamma0 - Teppeki",
                                          "gamma1[1]" = "gamma1 - untreated", "gamma1[2]" = "gamma1 - Mavrik Jet",
                                         "gamma1[3]" = "gamma1 - Movento", "gamma1[4]" = "gamma1 - Teppeki"  )) %>%
    filter(!grepl("Eff", Parameter)) %>%

@@

tab4 %>% 
  kbl(booktabs = TRUE, caption = "Summary of the posterior distributions obtained with the 100\\% Y - 0\\% Z scenario for the model parameters and treatment efficacies: posterior mean, standard deviation, 2.5 and 97.5 quantiles.") %>%
  kable_styling()
```

@@

```{r figure4, cache = FALSE, echo=FALSE, out.width = "100%", fig.height = 2.6}
#| label: fig-four
#| fig-cap: "Estimated treatment efficacies after 6 days (**A**) and after 12 days (**B**), with their credibility intervals. Colors correspond to the different scenarios."

@@

```{r table5, cache = FALSE}
#|echo: false
cbind(df_len6, df_len12 %>% select(- Data, - Insecticide)) %>% 
  kbl(booktabs = TRUE, caption = "Differences in the sizes of the 95 credibility intervals (CI) of the estimated treatment efficacies for the scenarios 50\\% Y - 50\\% Z, 0\\% Y - 50\\% Z and 100\\% Y - 0\\% Z, compared to 50\\% Y - 0\\% Z. The difference is given in percentage. A positive (negative) value indicates an increase (decrease) of the credibility interval size. The third column indicates differences for the efficacy at 6 days after pesticide spray, and the fourth column indicates the difference for the efficacy at 12 days.") %>% 
  kable_styling()
```

@@

```{r load_res_simulations, cache = FALSE, echo = FALSE}
df_col_simu = readRDS(file = "Files_for_code/Results/Simulations/df_col_simu.rds")
res_simu_sigma_1.87 = readRDS(file = "Files_for_code/Results/Simulations/res_simu_sigma_1.87.rds")

@@

```{r table6, cache = FALSE, echo = FALSE}

@@

@@

```{r figure5, cache = FALSE, out.width="100%", fig.height = 2.5, echo = FALSE}
#| label: fig-five
#| fig-cap: "Values of the $E_{\\gamma}$ (@eq-E_gamma) (**A**), $E_{Ef_6}$ (@eq-E_efficacy) (**B**) and $E_{Ef_{12}}$ (@eq-E_efficacy) (**C**) mean absolute error criteria for the '0% Y - 50% Z', '50% Y - 0% Z' and '50% Y - 50% Z' scenarios. The x-axis presents the number of trials and the y-axis presents the mean absolute error, averaged over the 974 simulated datasets. Different colors correspond to different scenarios."

@@

```{r figure6, cache = FALSE, out.width="100%", fig.height = 3, echo = FALSE}
#| label: fig-six
#| fig-cap: "Comparison of proportion cases where the best treatment is correctly identified in the '0% Y - 50% Z', '50% Y - 0% Z' and '50% Y - 50% Z' scenarios. The x-axis represents the number of trials and the y-axis represents the percentage of cases where the best treatment has been correctly identified at 6 days (A) and 12 days (B), over the 974 simulated datasets. Different colors correspond to different scenarios."

@@

```{r figure7, cache = FALSE, out.width="100%", fig.height = 2.5, echo = FALSE}
#| label: fig-seven
#| fig-cap: "Values of $E_{\\gamma}$ (@eq-E_gamma) (**A**), $E_{Ef_6}$ (@eq-E_efficacy) (**B**) and $E_{Ef_{12}}$ (@eq-E_efficacy) (**C**) for the '0% Y - 100% Z', '50% Y - 0% Z', '50% Y - 100% Z' and '50% Y - 50% Z' scenarios. The x-axis presents the number of trials and the y-axis presents the absolute error averaged over the 974 simulated datasets. Different colors correspond to different scenarios."

@@

```{r table7, cache = FALSE}
#|echo: false

@@

```{r figure8, cache = FALSE, out.width="100%", fig.height = 6, echo = FALSE}
#| label: fig-height
#| fig-cap: "Comparison of the '100% Y - 0% Z', '0% Y - 100% Z', '100% Y / 100% Z' and '100% W' scenarios according to the distribution of $Z$ and the number of trials, using the $E_{\\gamma}$ criterion (@eq-E_gamma). **A**, **B** and **C** correspond to different $Z$ distributions which are given by A.2, B.2 and C.2 (distribution for a number of trials equal to 40). A, B and C respectively correspond to $\\alpha_0 =$ -1, 1 and 2. The details of the simulation parameters are given in table 7. A1, B1 and C1 represent the absolute error $E_{\\gamma}$ averaged over the 974 simulated datasets as a function of the number of trials. Colors correspond to the different scenarios."
